Reconstructing contact network parameters from viral phylogenies

### 2016-07-14 doc/paper/changes.tex
Finished reviewer things finally ugh

Although they are very approximate methods, our experience has been that using
approximate maximum likelihood phylogeny reconstruction with root-to-tip
regression does not result in large errors in phylodynamic inference (Poon
2015). We reanalysed two datasets (Niculescu et al. 2015 and Cuevas et al.
2009) using RAxML and LSD. The results for $\alpha$ were nearly identical, but
$I$ was not robust to the reconstruction method -- RAxML+LSD produced higher
$I$ estimates for both datasets (Figure S13). This result has been added to the
discussion of both parameters.

Rather than analyze bootstrap replicates, we used the two new datasets
containing both \textit{gag} and \textit{env} sequences for within-dataset
replication.

>>>>

Although they are very approximate methods, our experience has been that using
approximate maximum likelihood phylogeny reconstruction with root-to-tip
regression does not result in large errors in phylodynamic inference (Poon
2015). We reanalysed the two datasets with the highest and lowest estimated
$\alpha$ values (Zetterberg et al. 2004 and Cuevas et al. 2009) using RAxML and
LSD. The results for $\alpha$ were quite similar, but $I$ was less robust to
the reconstruction method -- RAxML+LSD produced higher $I$ estimates for both
datasets (Figure S13). This result has been added to the discussion of both
parameters.

We also analyzed five bootstrap replicate alignments for each of these
datasets (Figure S15). The posterior means for the replicates were in the same
region as the posterior mean for the original alignment. For $\alpha$, the
means spanned a region of width approximately 0.3 (about 15\% of the prior
region); for $I$, the range was only about 5\% of the prior region. Two new
datasets containing both \textit{gag} and \textit{env} sequences provide
additional within-dataset replication.

### 2016-07-14 doc/paper/paper.Rnw
Finished reviewer things finally ugh

\add{To check if our results were robust to the choice of phylogenetic
reconstruction method, we built and reanalyzed phylogenies for two of the
datasets (mixed/Spain and IDU/Romania) with
\textit{RAxML}~\autocite{stamatakis2014raxml} with the GTR+$\Gamma$ model of
sequence evolution and rate heterogeneity. The trees were rooted and
time-scaled with \textit{Least Square
Dating}~\autocite[\textit{LSD},][]{to2016fast}.}

@@

For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
value $m = 1$ produces networks with no cycles, which we considered fairly
implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
the SMC algorithm were the same as used for the simulation experiments,
\add{except that we used 10000 particles instead of 1000 to increase the
accuracy of the estimated posterior. This was computationally feasible due to
the small number of runs required for this analysis.}

@@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ were apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
when all other parameters were fixed. \Cref{fig:better} shows the impact of
increasing the number of particles, simulated datasets, and
$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
number of iterations until the stopping condition was reached was 81 with the
basic settings and 124 with the higher settings. The results of the two
simulations were very similar for all parameters.}

@@

  stopifnot(pmean1[,dataset[which.min(I)]] == "kao2011")
  stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "li2015")

@@

\add{Posterior mean point estimates for the preferential attachment power $\alpha$
were all sub-linear, ranging from 
    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
(mixed/Spain) to
    \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
(IDU/Estonia). When aggregated by risk group, the average estimates were
    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015"), round(mean(alpha), 2)]}
for IDU,
    \Sexpr{pmean1[c("novitsky2014", "mccormack2002", "grabowski2014"), round(mean(alpha), 2)]}
for primarily heterosexual risk, and
    \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
for MSM. 95\% HPD intervals were very wide for most datasets, often
encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}).}

@@

\add{For all but the HET/Botswana data, the posterior mean estimates for
$I$ were between
    \Sexpr{pmean1["kao2011", round(I)]}
(MSM/Taiwan) and
    \Sexpr{pmean1["li2015", round(I)]}
(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
    (\Sexpr{pmean1["novitsky2014", round(I)]})
than the other datasets, with a very wide 95\% HPD interval covering almost the
entire prior region (\cref{fig:abchpd}). There was no significant correlation
between the number of sequences in the tree and the estimated $I$ value
(Spearman correlation, 
    $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
indicating that the higher estimates were not simply due to increased sampling
density. When both \textit{gag} and \textit{env} sequences were analyzed, the
estimates from the \textit{env} data were higher (HET/Uganda,
    \Sexpr{pmean1["grabowski2014", round(I)]} for \textit{gag} vs.
    \Sexpr{pmean1.env["grabowski2014", round(I)]} for \textit{env};
HET/Malawi,
    \Sexpr{pmean1["mccormack2002", round(I)]} for \textit{gag} vs.
    \Sexpr{pmean1.env["mccormack2002", round(I)]} for \textit{env}).}

@@

\add{When the value $m = 1$ was disallowed by the prior, the separation in
$\alpha$ between the IDU datasets and the others became more striking
(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
or above
    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015"), round(min(alpha), 2)]}.
The estimate for the MSM/Beijing data was slightly lower
    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
the estimates for the seven remaining non-IDU datasets were bounded above by
    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}. 
The values of $I$ were fairly robust to the choice of prior (compare
\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
narrower (average width
    \Sexpr{round(mean(up1[,I] - low1[,I]))}
for $m \geq 1$ and
    \Sexpr{round(mean(up2[,I] - low2[,I]))}
for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
took on the value 3 with this prior, with the HPD intervals spanning the entire
prior region. This is very similar to the results observed for $m$ on simulated
data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
data with this prior. The results for $N$ did not change appreciably between
the two choices of prior.}

@@

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics. When
interpreting these results, we caution that the BA model is quite simple and
most likely misspecified for these data. In particular, the average degree of a
node in the network is equal to $2m$, and therefore is constrained to be a
multiple of 2. Furthermore, we considered the case $m = 1$, where the network
has no cycles, to be implausible and therefore assigned it zero prior
probability in one set of analyses. This forced the average degree to be at
least four, which may be unrealistically high for sexual networks. The fact
that the estimated values of $\alpha$ differed substantially for three datasets
depending on whether or not $m = 1$ was allowed by the prior is further evidence
of this potential misspecification. However, we note that \add{the ordering of
the datasets with respect to $\alpha$ was similar between the two priors,}
\del{for two of the datasets, the estimated values of $\alpha$ did not change
much between priors,} and the estimates of $I$ were robust to the choice of
prior for all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
sophisticated models, for example models incorporating heterogeneity in node
behaviour, are likely to provide a better fit to these data.

@@

\add{The results we have just discussed pertain to the \textit{gag} sequences
for the two datasets with both \textit{gag} and \textit{env} data. As shown in
\cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
analyzed. For the two datasets we reanalyzed using
RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
was robust to the choice of method (\cref{fig:methods}).}

@@

<<realdata_I, include=FALSE>>=
    #stopifnot(d1[parameter == "I", dataset[which.min(mean)]] == "zetterberg2004")
    stopifnot(d1[parameter == "I", dataset[which.max(mean)]] == "novitsky2014")
@

\add{The true HIV prevalence in a population can be difficult to estimate for
several reasons. HIV-infected individuals may be asymptomatic for months or
years, possibly delaying their awareness of their status. In many contexts, the
risk factors for acquisition of HIV are illegal or stigmatized, which may
represent a barrier to testing, treatment, and/or disclosure of status.
Our simulation study showed that $I$ is weakly identifiable from tree shapes,
however, the estimates of $I$ obtained with \textit{netabc} were upwardly
biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
that the identifiability of $I$ decreases with the number of sampled tips
(\cref{fig:Icrossv}); in real world studies, the proportion of infected
individuals sampled is usually low. The estimated $I$ values for the HIV
datasets ranged from
    \Sexpr{d1[parameter == "I", round(min(mean))]}
    (IDU/Estonia)
to
    \Sexpr{d1[parameter == "I", round(max(mean))]}
    (HET/Botswana).
The estimates of $I$ were less robust to the HIV gene analyzed
(\cref{fig:genes}) and the phylogenetic reconstruction method used
(\cref{fig:methods}).}

@@

Our method has a number of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating} due to within-host evolution and the
sampling process. \add{Progress has very recently been made on this front by
\textcite{giardina2016inference}, who have concurrently developed an ABC-based
method incorporating within-host evolution.} The ABC-SMC algorithm is
computationally intensive, taking about a day when run on 20 cores in parallel
with the settings we described in the methods. Nevertheless, our method is
potentially useful to epidemiological researchers interested in the general
characteristics of the network structure underlying disease outbreaks. This
work, and previous work by our group~\autocite{poon2015phylodynamic}, has
demonstrated that ABC is a broadly applicable and effective framework in which
to perform phylodynamic inference.

@@

\clearpage
\newpage
\printbibliography

@@

\begin{figure}[ht]
  \centering
  \includegraphics{derivatives.pdf}
  \caption[
    First and second time derivatives of epidemic growth curves at time of
    sampling for various values of $I$ and $N$.
  ]{
    First and second \add{time} derivatives of epidemic growth curves at time
    of sampling for various values of $I$ and $I$. Networks were simulated
    under the BA model with $\alpha = 1.0$, $m = 2$, and $N$ varied along the
    values shown on the $x$-axis. Transmission trees were sampled at the time
    when $I$ nodes were infected ($y$-axis). Logistic growth curves were fit to
    epidemic trajectories derived from the transmission trees, and their first
    and second derivatives were calculated at the time of sampling. Contours
    show first derivatives, while colours indicate second derivatives. Values
    of $I$ and $N$ used in simulation experiments with ABC are indicated by
    diamonds.
  }
  \label{fig:derivs}
\end{figure}

>>>>

\add{To check if our results were robust to the choice of phylogenetic
reconstruction method, we built and reanalyzed phylogenies for the datasets
with the lowest and highest estimated $\alpha$ values (mixed/Spain and
IDU/Estonia) with \textit{RAxML}~\autocite{stamatakis2014raxml} with the
GTR+$\Gamma$ model of sequence evolution and rate heterogeneity. The trees were
rooted and time-scaled with \textit{Least Square
Dating}~\autocite[\textit{LSD},][]{to2016fast}. For expediency, the analysis
was run with the prior $m \sim \text{DiscreteUniform}(2, 5)$, which defines a
smaller total search space than the prior allowing $m = 1$. For both of these
datasets, we also analyzed five bootstrap replicate alignments generated by
resampling alignment columns with replacement.}

@@

For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
value $m = 1$ produces networks with no cycles, which we considered fairly
implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
the SMC algorithm were the same as used for the simulation experiments,
\add{except that we used 10000 particles instead of 1000 to increase the
accuracy of the estimated posterior for all analyses except the bootstrap
replicates. This was computationally feasible due to the small number of runs
required for this analysis.}

@@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ were apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
when all other parameters were fixed. \Cref{fig:better} shows the impact of
increasing the number of particles, simulated datasets, and
$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
number of iterations until the stopping condition was reached was 81 with the
basic settings and 124 with the higher settings. The results of the two
simulations were similar, but surprisingly, the results with higher SMC
settings were slightly worse (by
    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
$I$ was closer to the true value of 2000 with the improved settings
    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
The estimate of $m$, 3 in both cases, was unaffected by the settings.}

@@

  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
  stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "li2015")

@@

<<raxml, include=FALSE>>=
    d.raxml <- fread(Sys.glob("../../simulations/aggregates/raxml-hpd/*"))
@

\add{Posterior mean point estimates for the preferential attachment power $\alpha$
were all sub-linear, ranging from 
    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
(mixed/Spain) to
    \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
(IDU/Estonia). When aggregated by risk group, the average estimates were
    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015"), round(mean(alpha), 2)]}
for IDU,
    \Sexpr{pmean1[c("novitsky2014", "mccormack2002", "grabowski2014"), round(mean(alpha), 2)]}
for primarily heterosexual risk, and
    \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
for MSM. 95\% HPD intervals were very wide for most datasets, often
encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}). As shown
in \cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
analyzed.}

@@

\add{For all but the HET/Botswana data, the posterior mean estimates for
$I$ were between
    \Sexpr{pmean1["zetterberg2004", round(I)]}
(IDU/Estonia) and
    \Sexpr{pmean1["li2015", round(I)]}
(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
    (\Sexpr{pmean1["novitsky2014", round(I)]})
than the other datasets, with a very wide 95\% HPD interval covering almost the
entire prior region (\cref{fig:abchpd}). There was no significant correlation
between the number of sequences in the tree and the estimated $I$ value
(Spearman correlation, 
    $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
indicating that the higher estimates were not simply due to increased sampling
density. When both \textit{gag} and \textit{env} sequences were analyzed, the
estimates from the \textit{env} data were higher (HET/Uganda,
    \Sexpr{pmean1["grabowski2014", round(I)]} for \textit{gag} vs.
    \Sexpr{pmean1.env["grabowski2014", round(I)]} for \textit{env};
HET/Malawi,
    \Sexpr{pmean1["mccormack2002", round(I)]} for \textit{gag} vs.
    \Sexpr{pmean1.env["mccormack2002", round(I)]} for \textit{env}).}

@@

\add{When the value $m = 1$ was disallowed by the prior, the separation in
$\alpha$ between the IDU datasets and the others became more striking
(\cref{fig:abchpdm2}). Both IDU datasets had estimated $\alpha$ values at or
above
    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015"), round(min(alpha), 2)]}.
The estimate for the MSM/Beijing data was slightly lower
    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
the estimates for the seven remaining non-IDU datasets were bounded above by
    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}. 
The values of $I$ were fairly robust to the choice of prior (compare
\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
narrower (average width
    \Sexpr{round(mean(up1[,I] - low1[,I]))}
for $m \geq 1$ and
    \Sexpr{round(mean(up2[,I] - low2[,I]))}
for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
took on the value 3 with this prior, with the HPD intervals spanning the entire
prior region. This is very similar to the results observed for $m$ on simulated
data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
data with this prior. The results for $N$ did not change appreciably between
the two choices of prior.}

<<bootstrap, include=FALSE>>=
    d.boot <- fread(Sys.glob("../../simulations/aggregates/bootstrap-hpd/*"))
@

\add{For the two datasets we reanalyzed using
RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
was relatively robust to the choice of method (\cref{fig:methods}, posterior
means 
    \Sexpr{pmean2["cuevas2009", round(alpha, 2)]}
vs. 
    \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "alpha", round(mean, 2)]}
for mixed/Spain and 
    \Sexpr{pmean2["zetterberg2004", round(alpha, 2)]}
vs.
    \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "alpha", round(mean, 2)]}
for IDU/Estonia). However, the estimates of $I$ were about twice as high when
RAxML was used instead of FastTree to reconstruct the trees 
    (\Sexpr{pmean2["zetterberg2004", round(I)]} 
     vs. \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "I", round(mean)]} for IDU/Estonia, 
    \Sexpr{pmean2["cuevas2009", round(I)]}
    vs. \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "I", round(mean)]} for mixed/Spain). 
\Cref{fig:boot} shows estimates obtained for five bootstrap replicate
alignments for each of these two datasets. For the mixed/Spain data, the
estimated posterior mean [range of bootstrap posterior means] was
    \Sexpr{pmean2["cuevas2009", round(alpha, 2)]}
    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "alpha" & replicate < 5, round(min(mean), 2)]} -
     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "alpha" & replicate < 5, round(max(mean), 2)]}]
for $\alpha$, 
    \Sexpr{pmean2["cuevas2009", round(I)]}
    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "I" & replicate < 5, round(min(mean))]} -
     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "I" & replicate < 5, round(max(mean))]}] for $I$,
    \Sexpr{pmean2["cuevas2009", round(m, 2)]}
    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "m" & replicate < 5, round(min(mean), 2)]} -
     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "m" & replicate < 5, round(max(mean), 2)]}] for $m$, and
    \Sexpr{pmean2["cuevas2009", round(N)]}
    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "N" & replicate < 5, round(min(mean))]} -
     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "N" & replicate < 5, round(max(mean))]}] for $N$. 
For the IDU/Estonia data, values were
    \Sexpr{pmean2["zetterberg2004", round(alpha, 2)]}
    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "alpha" & replicate < 5, round(min(mean), 2)]} -
     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "alpha" & replicate < 5, round(max(mean), 2)]}]
for $\alpha$, 
    \Sexpr{pmean2["zetterberg2004", round(I)]}
    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "I" & replicate < 5, round(min(mean))]} -
     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "I" & replicate < 5, round(max(mean))]}] for $I$,
    \Sexpr{pmean2["zetterberg2004", round(m, 2)]}
    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "m" & replicate < 5, round(min(mean), 2)]} -
     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "m" & replicate < 5, round(max(mean), 2)]}] for $m$, and
    \Sexpr{pmean2["zetterberg2004", round(N)]}
    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "N" & replicate < 5, round(min(mean))]} -
    \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "N" & replicate < 5, round(max(mean))]}] for $N$.}

@@

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics. When
interpreting these results, we caution that the BA model is quite simple and
most likely misspecified for these data. In particular, the average degree of a
node in the network is equal to $2m$, and therefore is constrained to be a
multiple of 2. Furthermore, we considered the case $m = 1$, where the network
has no cycles, to be implausible and therefore assigned it zero prior
probability in one set of analyses. This forced the average degree to be at
least four, which may be unrealistically high for sexual networks. The fact
that the estimated values of $\alpha$ differed substantially for several datasets
depending on whether or not $m = 1$ was allowed by the prior is further evidence
of this potential misspecification. However, we note that \add{the ordering of
the datasets with respect to $\alpha$ was similar between the two priors,}
\del{for two of the datasets, the estimated values of $\alpha$ did not change
much between priors,} and the estimates of $I$ were robust to the choice of
prior for all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
sophisticated models, for example models incorporating heterogeneity in node
behaviour, are likely to provide a better fit to these data.

@@

@@

<<realdata_I, include=FALSE>>=
    stopifnot(d1[parameter == "I", dataset[which.min(mean)]] == "zetterberg2004")
    stopifnot(d1[parameter == "I", dataset[which.max(mean)]] == "novitsky2014")
@

\add{The true HIV prevalence in a population can be difficult to estimate for
several reasons. HIV-infected individuals may be asymptomatic for months or
years, possibly delaying their awareness of their status. In many contexts, the
risk factors for acquisition of HIV are illegal or stigmatized, which may
represent a barrier to testing, treatment, and/or disclosure of status.
Our simulation study showed that $I$ is weakly identifiable from tree shapes,
however, the estimates of $I$ obtained with \textit{netabc} were upwardly
biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
that the identifiability of $I$ decreases with the number of sampled tips
(\cref{fig:Icrossv}); in real world studies, the proportion of infected
individuals sampled is usually low. The estimated $I$ values for the HIV
datasets ranged from
    \Sexpr{d1[parameter == "I", round(min(mean))]}
    (IDU/Estonia)
to
    \Sexpr{d1[parameter == "I", round(max(mean))]}
(HET/Botswana).}

@@

Our method has a number of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating} due to within-host evolution and the
sampling process~\autocite{giardina2016inference}. The ABC-SMC algorithm is
computationally intensive, taking about a day when run on 20 cores in parallel
with the settings we described in the methods. Nevertheless, our method is
potentially useful to epidemiological researchers interested in the general
characteristics of the network structure underlying disease outbreaks. This
work, and previous work by our group~\autocite{poon2015phylodynamic}, has
demonstrated that ABC is a broadly applicable and effective framework in which
to perform phylodynamic inference.

@@

\clearpage
\printbibliography
\newpage

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{realdata-bootstrap}
    \caption[
        Posterior means and 50\%/95\% HPD intervals for parameters of the
        BA network model, fitted to original and bootstrap replicate alignments
        of two HIV datasets.
    ]{
        Posterior means (points), 50\% HPD intervals (notches), and 95\%
        HPD intervals (lines) for parameters of the BA network model, fitted to
        original and bootstrap replicate alignments of two HIV datasets.
    }
    \label{fig:boot}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{derivatives.pdf}
  \caption[
    First and second time derivatives of epidemic growth curves at time of
    sampling for various values of $I$ and $N$.
  ]{
    First and second time derivatives of epidemic growth curves at time
    of sampling for various values of $I$ and $I$. Networks were simulated
    under the BA model with $\alpha = 1.0$, $m = 2$, and $N$ varied along the
    values shown on the $x$-axis. Transmission trees were sampled at the time
    when $I$ nodes were infected ($y$-axis). Logistic growth curves were fit to
    epidemic trajectories derived from the transmission trees, and their first
    and second derivatives were calculated at the time of sampling. Contours
    show first derivatives, while colours indicate second derivatives. Values
    of $I$ and $N$ used in simulation experiments with ABC are indicated by
    diamonds.
  }
  \label{fig:derivs}
\end{figure}
